\pkg{casebase}: An Alternative Framework For Survival Analysis

### 2020-02-06 analysis/paper/paper.tex
Switched to non interaction model
note that I only ran CS3 so the pdf is not complete. fixes include all red text

Jesse Still needs to clean up the text and deliberate his write up in the case study. Code will also be cleaned to be more readable.

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \usepackage{tabu} \DeclareUnicodeCharacter{2500}{-}

@@

{\centering \includegraphics{../figures/erspc-cox-cif-1}

\caption[Cumulative incidence functions for control and screening groups in the ERSPC data]{Cumulative incidence functions for control and screening groups in the ERSPC data. In each of the panels, we plot the CIF from the Cox model (solid line) and the CIF from the caseabse sampling scheme (dashed line) with different functional forms of time in a logistic regression. (A) The time variable is excluded so that the hazard function is constant over time which is equivalent to the one-parameter exponential distribution. (B) Linear function of time which is equivalent to a Gompertz distribution. (C) The natural logarithm of time which is equivalent to a Weibull distribution. (D) Cubic b-spline expansion of time.}\label{fig:erspc-cox-cif}
\end{figure}
\end{CodeChunk}

@@

\begin{center}\includegraphics{../figures/unnamed-chunk-1-1} \end{center}

\hypertarget{equivalence-between-casebase-and-aft}{%
\subsection{Equivalence between casebase and
AFT}\label{equivalence-between-casebase-and-aft}}

@@

\begin{CodeChunk}
\begin{table}

\caption{\label{tab:unnamed-chunk-4}Mean }
\centering
\begin{tabular}[t]{>{\bfseries}lcc}
\toprule
Model & casebase & survreg\\
\midrule
Exponential & 0.81 (0.68, 0.96) & 0.81 (0.68, 0.96)\\
\cmidrule{1-3}
Gompertz & 0.78 (0.66, 0.93) & 0.80 (0.67, 0.95)\\
\cmidrule{1-3}
Weibull & 0.80 (0.67, 0.95) & 0.80 (0.65, 0.96)\\
\cmidrule{1-3}
Splines & 0.81 (0.68, 0.96) & --\\
\bottomrule
\multicolumn{3}{l}{\textit{Note:}}\\
\multicolumn{3}{l}{Median (Inter-quartile range) is given for Model Size.}\\
\end{tabular}
\end{table}

>>>>

\usepackage{amsmath} \usepackage{longtable} \usepackage{graphicx} \usepackage{tabularx} \usepackage{float} \usepackage{booktabs} \usepackage{makecell} \DeclareUnicodeCharacter{2500}{-}

@@

@@

\begin{CodeInput}
R> new_data <- data.frame(ScrArm = c("Control group", "Screening group"),
R>                        ignore = 99)
R> 
R> plot(survfit(cox_model, newdata = new_data),
R>      xlab = "Years since Randomization", 
R>      ylab = "Cumulative Incidence (%)", 
R>      fun = "event",
R>      xlim = c(0,15), conf.int = FALSE, col = c("red","blue"), 
R>      main = sprintf("Estimated Cumulative Incidence (risk) of Death from Prostate 
R>                     Cancer Screening group Hazard Ratio: %.2g (%.2g, %.2g)",
R>                     exp(coef(cox_model)), 
R>                     exp(confint(cox_model))[1], 
R>                     exp(confint(cox_model))[2]),
R>      ylim = c(0, 2), yscale = 100)
R> legend("topleft", 
R>        legend = c("Control group", "Screening group"), 
R>        col = c("red","blue"),
R>        lty = c(1, 1), 
R>        bg = "gray90")
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeInput}
R> anova(casebase_exponential, casebase_splines, test = "LRT")
\end{CodeInput}
\end{CodeChunk}

As expected, we see that splines model provides a better fit. Similarly,
we can compare the AIC for all three case-base models:

\begin{CodeInput}
R> c("Exp." = AIC(casebase_exponential),
R>   "Weib." = AIC(casebase_time),
R>   "Splines" = AIC(casebase_splines))
\end{CodeInput}
\end{CodeChunk}

Once again, we have evidence that the splines model provides the best
fit.

With this first case study, we saw how population-time plots can provide
a useful summary of the incidence for our event of interests. In
particular, we saw how it can provide evidence against an exponential
hazard model. We then explored how different parametric models can
provide different estimates of the CIF, even though they all give
similar estimates for the hazard ratios.

\hypertarget{case-study-2bone-marrow-transplant}{%
\section{Case study 2--Bone-marrow
transplant}\label{case-study-2bone-marrow-transplant}}

In the next case study, we show how case-base sampling can also be used
in the context of a competing risk analysis. For illustrative purposes,
we will use the same data that was used in Scrucca \emph{et al}
\citeyearpar{scrucca2010regression}. The data was downloaded from the
first author's website, and it is also available as part of the
\pkg{casebase} package.

\begin{CodeInput}
R> 
R> data(bmtcrr)
\end{CodeInput}
\end{CodeChunk}

The data contains information on 177 patients who received a stem-cell
transplant for acute leukemia. The event of interest is relapse, but
other competing causes (e.g.~transplant-related death) were also
recorded. Several covariates were captured at baseline: sex, disease
type (acute lymphoblastic or myeloblastic leukemia, abbreviated as ALL
and AML, respectively), disease phase at transplant (Relapse, CR1, CR2,
CR3), source of stem cells (bone marrow and peripheral blood, coded as
BM+PB, or only peripheral blood, coded as PB), and age. A summary of
these baseline characteristics appear in Table \ref{tab:table1bmtcrr}.
We note that the statistical summaries were generated differently for
different variable types: for continuous variables, we gave the range,
followed by the mean and standard deviation; for categorical variables,
we gave the counts for each category.

First, we can look at a population-time plot to visualize the incidence
density of both relapse and the competing events. In Figure
\ref{fig:compPop}, failure times are highlighted on the plot using red
points for the event of interest (panel A) and blue dots for competing
events (panel B). In both panels, we see evidence of a non-constant
hazard function: the density of points is larger at the beginning of
follow-up than at the end.

Our main objective is to compute the absolute risk of relapse for a
given set of covariates. First, we fit a smooth hazard to the data using
a linear term for time:

\begin{CodeInput}
R> model_cb <- fitSmoothHazard(
R>     Status ~ ftime + Sex + D + Phase + Source + Age, 
R>     data = bmtcrr, 
R>     ratio = 100, 
R>     time = "ftime")
\end{CodeInput}
\end{CodeChunk}

We will compare our hazard ratio estimates to that obtained from a Cox
regression. To do so, we need to treat the competing event as censoring.

\begin{CodeInput}
R> library(survival)
R> # Treat competing event as censoring
R> model_cox <- coxph(Surv(ftime, Status == 1) ~ Sex + D + Phase + Source + Age,
R>                    data = bmtcrr)
\end{CodeInput}
\end{CodeChunk}

From the fit object, we can extract both the hazard ratios and their
corresponding confidence intervals. These quantities appear in Table
\ref{tab:bmtcrr-cis}. As we can see, the only significant hazard ratio
identified by case-base sampling is the one associated with the phase of
the disease at transplant. More precisely, being in relapse at
transplant is associated with a hazard ratio of 3.89 when compared to
CR1.

Given the estimate of the hazard function obtained using case-base
sampling, we can compute the absolute risk curve for a fixed covariate
profile. We perform this computation for a 35 year old woman who
received a stem-cell transplant from peripheral blood at relapse. We
compared the absolute risk curve for such a woman with acute
lymphoblastic leukemia with that for a similar woman with acute
myeloblastic leukemia. We will estimate the curve from 0 to 60 months.

\begin{CodeInput}
R> # Pick 100 equidistant points between 0 and 60 months
R> time_points <- seq(0, 60, length.out = 50)
R> 
R> # Data.frame containing risk profile
R> newdata <- data.frame("Sex" = factor(c("F", "F"), 
R>                                      levels = levels(bmtcrr[,"Sex"])),
R>                       "D" = c("ALL", "AML"), # Both diseases
R>                       "Phase" = factor(c("Relapse", "Relapse"), 
R>                                        levels = levels(bmtcrr[,"Phase"])),
R>                       "Age" = c(35, 35),
R>                       "Source" = factor(c("PB", "PB"), 
R>                                         levels = levels(bmtcrr[,"Source"])))
R> 
R> # Estimate absolute risk curve
R> risk_cb <- absoluteRisk(object = model_cb, time = time_points,
R>                         method = "numerical", newdata = newdata)
\end{CodeInput}
\end{CodeChunk}

We will compare our estimates to that obtained from a corresponding
Fine-Gray model \citep{fine1999proportional}. The Fine-Gray model is a
semiparametric model for the cause-specific \emph{subdistribution},
i.e.~the function \(f_k(t)\) such that
\[CI_k(t) =1 - \exp\left( - \int_0^t f_k(u) \textrm{d}u \right),\] where
\(CI_k(t)\) is the cause-specific cumulative indicence. The Fine-Gray
model allows to directly assess the effect of a covariate on the
subdistribution, as opposed to the hazard. For the computation, we will
use the \pkg{timereg} package \citep{timereg}.

\begin{CodeInput}
R> library(timereg)
R> model_fg <- comp.risk(Event(ftime, Status) ~ const(Sex) + const(D) + 
R>                           const(Phase) + const(Source) + const(Age), 
R>                       data = bmtcrr, cause = 1, model = "fg")
R> 
R> # Estimate absolute risk curve
R> risk_fg <- predict(model_fg, newdata, times = time_points)
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:compAbsrisk} shows the absolute risk curves for both
case-base sampling and the Fine-Gray model. As we can see, the two
approaches agree quite well for acute myeloblastic leukemia; however,
there seems to be a difference of about 5\% between the two curves for
acute lymphoblastic leukemia. This difference does not appear to be
significant: the curve from case-base sampling is contained within a
95\% confidence band around the Fine-Gray absolute risk curve (figure
not shown).

\hypertarget{case-study-3support-data}{%
\section{Case study 3--SUPPORT Data}\label{case-study-3support-data}}

In the first two case studies, we described the basic functionalities of
the \pkg{casebase} package: creating population-time plots, fitting
parametric models for hazard functions, and estimating the corresponding
cumulative incidence curves. In the next two case studies, we explore
more advanced features.

For the third case study, we will introduce regularization into the
estimation of the hazard function. This can be performed using a
combination of case-base sampling and regularized logistic regression.
To demonstrate this capability, we examine the Study to Understand
Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) dataset
\citep{knaus1995support}. The SUPPORT dataset tracks death for
individuals who are considered seriously ill within a hospital.
Imputation was conducted using the \pkg{mice} package in \proglang{R}
with default settings \citep{mice}. Before imputation, there were a
total of 1000 individuals and 35 variables. charges, Totcst, totmcst and
sfdm2 were removed as they are response variables we are not interested
in for the sake of this analysis. After imputation, we have 690
individuals. We lose individuals due to coliniarity between variables,
preventing imputation. If a pair of covariates are deemed colinear to a
certain degree, they are excluded from any imputations. Here, only edu
income avtisst* race ph* glucose* adlp* were imputed. Of these 690
individuals, 66.09\% experienced the event of interest. A description of
each variable can be found in Table \ref{tab:support1} and a breakdown
of each categorical variable in Table \ref{tab:support2}. The original
data was obtained from the Department of Biostatistics at Vanderbilt
University. The imputed dataset is available as part of the
\pkg{casebase} package.
\begin{tabular}{ccccc}
\textbf{Name} & \textbf{Labels}                          & \textbf{Levels} & \textbf{Storage} & \textbf{NAs} \\
age           & Age                                      &                 & double           & 0            \\
death         & Death at any time up to NDI date:31DEC94 &                 & double           & 0            \\
sex           &                                          & 2               & integer          & 0            \\
hospdead      & Death in Hospital                        &                 & double           & 0            \\
slos          & Days from Study Entry to Discharge       &                 & double           & 0            \\
d.time        & Days of Follow-Up                        &                 & double           & 0            \\
dzgroup       &                                          & 8               & integer          & 0            \\
dzclass       &                                          & 4               & integer          & 0            \\
num.co        & number of comorbidities                  &                 & double           & 0            \\
edu           & Years of Education                       &                 & double           & 202          \\
income        &                                          & 4               & integer          & 349          \\
scoma         & SUPPORT Coma Score based on Glasgow D3   &                 & double           & 0            \\
charges       & Hospital Charges                         &                 & double           & 25           \\
totcst        & Total RCC cost                           &                 & double           & 105          \\
totmcst       & Total micro-cost                         &                 & double           & 372          \\
avtisst       & Average TISS, Days 3-25                  &                 & double           & 6            \\
race          &                                          & 5               & integer          & 5            \\
meanbp        & Mean Arterial Blood Pressure Day 3       &                 & double           & 0            \\
wblc          & White Blood Cell Count Day 3             &                 & double           & 24           \\
hrt           & Heart Rate Day 3                         &                 & double           & 0            \\
resp          & Respiration Rate Day 3                   &                 & double           & 0            \\
temp          & Temperature (Celsius) Day 3              &                 & double           & 0            \\
pafi          & PaO2/(.01*FiO2) Day 3                    &                 & double           & 253          \\
alb           & Serum Albumin Day 3                      &                 & double           & 378          \\
bili          & Bilirubin Day 3                          &                 & double           & 297          \\
crea          & Serum creatinine Day 3                   &                 & double           & 3            \\
sod           & Serum sodium Day 3                       &                 & double           & 0            \\
ph            & Serum pH (arterial) Day 3                &                 & double           & 250          \\
glucose       & Glucose Day 3                            &                 & double           & 470          \\
bun           & BUN Day 3                                &                 & double           & 455          \\
urine         & Urine Output Day 3                       &                 & double           & 517          \\
adlp          & ADL Patient Day 3                        &                 & double           & 634          \\
adls          & ADL Surrogate Day 3                      &                 & double           & 310          \\
sfdm2         &                                          & 5               & integer          & 159          \\
adlsc         & Imputed ADL Calibrated to Surrogate      &                 & double           & 0           
\end{tabular}
\caption{A description of each variable in the SUPPORT dataset.}
\label{tab:support1}

\begin{table}[ht]
\centering
\begin{tabular}{cc}
\hline
\textbf{Variable} & \textbf{Levels}                      \\
\hline
sex      & female                                        \\
         & male                                          \\
dzgroup  & ARF/MOSF w/Sepsis                             \\
         & COPD                                          \\
         & CHF                                           \\
         & Cirrhosis                                     \\
         & Coma                                          \\
         & Colon Cancer                                  \\
         & Lung Cancer                                   \\
         & MOSF w/Malig                                  \\
dzclass  & ARF/MOSF                                      \\
         & COPD/CHF/Cirrhosis                            \\
         & Coma                                          \\
         & Cancer                                        \\
income   & under \$11k                                   \\
         & $11-$25k                                      \\
         & $25-$50k                                      \\
         & \textgreater{}\$50k                           \\
race     & white                                         \\
         & black                                         \\
         & asian                                         \\
         & other                                         \\
         & hispanic                                      \\
sfdm2    & no(M2 and SIP pres)                           \\
         & adl\textgreater{}=4 (\textgreater{}=5 if sur) \\
         & SIP\textgreater{}=30                          \\
         & Coma or Intub                                 \\
         & \textless{}2 mo. follow-up                   
\end{tabular}
\caption{A description of each level within each categorical variable in the dataset.}
\label{tab:support2}
\end{table}

As before, the first step is to visualize the incidence density. The
population-time plot for the SUPPORT data appears in Figure
\ref{fig:support-poptime}.

{\centering \includegraphics{../figures/support-poptime-1}

\caption{\label{fig:support-poptime} Population-time plot for the SUPPORT data. Red points represent the case series. The gray space demonstrates the base series.}\label{fig:support-poptime}
\end{figure}
\end{CodeChunk}

Before we proceed with the analysis, we first remove hospital death as a
covariate, as this is directly informative of death. All covariates will
be used to estimate the hazard function. The hazard function will be
used to estimate the cumulative incidence of each individual in the
study. Then, the average cumulative incidience curve for all
observations will be used for our comparison.

Our main objective is to compute the absolute risk of death for a given
set of covariates, using regularization when fitting our model. First,
we fit a smooth hazard to the data using a weibull distribution. As
casebase makes use of the glmnet package, we interact with the
fitsmoothhazard.fit function using a matrix interface, where y contains
the time and event variables and x contains all other variables we would
like to include in the model. For the SUPPORT dataset, all factor
variables were converted into indicator variables. Here, we used lasso
penalization by setting alpha to 1.

\begin{CodeInput}
R> 
R> cb.Model=casebase::fitSmoothHazard.fit(x,y,family="glmnet",time="d.time",event="death",
R+                                        formula_time = ~log(d.time),alpha=1,ratio=ratio)
\end{CodeInput}
\end{CodeChunk}

Then, we take our model and use the \texttt{absoluteRisk} function to
integrate and retrieve our cumulative incidence curves.

We will compare this curve to a regularized version of cox regression,
and a kaplan-meier survival curve. The cox regression absolute risk
curve with lasso penalization required two extra steps to retrieve.
Coxnet from the glmnet package has tools to fit a regularized cox model,
but requires a survival object skeleton of the selected variables, so
that the survival package tools can be used to retrive the absolute
risk. We handle this by fitting a regularized cox model with the glmnet
package, and a cox model from the survival package with the selected
variables from the regularized fit. This second model serves as a
skeleton that permits the use of survival package functions. The
coefficients from the regularized model will replace the coefficients in
our skeleton. the resulting model will have the correct coefficients
when integrating, but will have an incorrect standard error. For the
purposes of this case study, we are only interested in the absolute risk
curve itself and not the standard error.

We used the survival package to calculate cox regression's cumulative
incidience without penalization and the kaplan-meier curve for
incidience.

\begin{CodeInput}
R> 
R> coxFitup <- survival::coxph(Surv(time=V1,event=V2) ~ ., data = Cox)
R> abcoxFitup<-survival::survfit(coxFitup,newdata=Cox)
R> 
R> #creating a surv object to be used to fit an unadjusted absolute risk curve.
R> # (Kaplan-Meier risk curve)
R> km <- survival::Surv(time = completeData$d.time, event = completeData$death)
R> abKm<-survival::survfit(km~1,type='kaplan-meier',conf.type='log')
\end{CodeInput}
\end{CodeChunk}

Now that our four cumulative incidence estimations have been calculated,
we compare them all in Figure \ref{fig:abSC}. Both coxnet and casebase
decrease the absolute risk by the end of the study, in comparison to
kaplan-meier.

{\centering \includegraphics{../figures/abSupportComparison-1}

\caption{\label{fig:abSC} Compares regularized casebase in black, regularized cox in red and Kaplan meier in blue.}\label{fig:abSupportComparison}
\end{figure}
\end{CodeChunk}

\begin{CodeInput}
R> 
R> #get covariates excluding d.time
R> estimatescb=setDT(as.data.frame(coef(cb.Model)[-c(1,2),1]), keep.rownames = TRUE)[]
R> estimatescb$Model="casebase"
R> colnames(estimatescb)<-c("Coefs","Estimate","Model")
R> estimatescox=setDT(as.data.frame(coef(coxglmFit)[-c(1),1]), keep.rownames = TRUE)[]
R> estimatescox$Model="Cox"
R> colnames(estimatescox)<-c("Coefs","Estimate","Model")
R> estimates=rbind(estimatescb, estimatescox)
R> estimates$Estimate[estimates$Estimate==0]=NA# make it so unchosen covariates appear empty
R> library(ggstance)# for vertical dodge
R>   ggplot(estimates, aes(x = Estimate, y = Coefs, color = Model)) +
R>         geom_segment(aes(x = 0, y = Coefs, xend = Estimate, yend = Coefs),size=0.5 ) +
R>         geom_point(position=ggstance::position_dodgev(height=0.2))
R> 
\end{CodeInput}
\end{CodeChunk}

\hypertarget{case-study-4stanford-heart-transplant-data}{%
\section{Case study 4--Stanford Heart Transplant
Data}\label{case-study-4stanford-heart-transplant-data}}

Although the previous case studies provide a broad overview of the
capabilities of \pkg{casebase}, they all have two properties in common:

\begin{itemize}
  \item All covariates are fixed, i.e. they do not change over time;
  \item The estimated hazard functions satisfy the "proportional hazard" assumption.
\end{itemize}

Case-base sampling has been used in the literature to study vaccination
safety, where the exposure period was defined as the week following
vaccination \citep{saarela2015case}. Hence, the main covariate of
interest, i.e.~exposure to the vaccine, was changing over time. In this
context, case-base sampling offers an efficient alternative to nested
case-control designs or self-matching.

The purpose of the next case study is to show how \pkg{casebase} can
also be used for survival analysis problems that do not share the
properties above. To this end, we will use the Stanford Heart Transplant
Data \citep[\citet{crowley1977covariance}]{clark1971cardiac}; we will
use the version available in the \pkg{survival} package.

Recall the setting of this study: patients were admitted to the Stanford
program after meeting with their physician and determining that they
were unlikely to respond to other forms of treatment. After enrollment,
the program searches for a suitable donor for the patient, which can
take anywhere between a few days to almost a year. We are interested in
the effect of a heart transplant on survival; therefore, the patient is
considered exposed only after the transplant has occurred.

As before, we can look at the population-time plot for a graphical
summary of the event incidence. As we can see, most events occur early
during the follow-up period, and therefore we do not expect the hazard
to be constant.

\begin{CodeInput}
R> library(survival)
R> library(casebase)
R> 
R> stanford_popTime <- popTime(jasa, time = "futime", 
R>                             event = "fustat")
R> plot(stanford_popTime)
\end{CodeInput}
\end{CodeChunk}

Since the exposure is time-dependent, we need to manually define the
exposure variable \emph{after} case-base sampling and \emph{before}
fitting the hazard function. For this reason, we will use the
\texttt{sampleCaseBase} function directly.

\begin{CodeInput}
R> library(tidyverse)
R> library(lubridate)
R> 
R> cb_data <- sampleCaseBase(jasa, time = "futime", 
R>                           event = "fustat", ratio = 10)
\end{CodeInput}
\end{CodeChunk}

Next, we will compute the number of days from acceptance into the
program to transplant, and we use this variable to determine whether
each population-moment is exposed or not.

\begin{CodeInput}
R> # Define exposure variable
R> cb_data <- mutate(cb_data,
R>                   txtime = time_length(accept.dt %--% tx.date, 
R>                                        unit = "days"),
R>                   exposure = case_when(
R>                     is.na(txtime) ~ 0L,
R>                     txtime > futime ~ 0L,
R>                     txtime <= futime ~ 1L
R>                   ))
\end{CodeInput}
\end{CodeChunk}

Finally, we can fit the hazard using various linear predictors.

\begin{CodeInput}
R> library(splines)
R> # Fit several models
R> fit1 <- fitSmoothHazard(fustat ~ exposure,
R>                         data = cb_data, time = "futime")
R> fit2 <- fitSmoothHazard(fustat ~ exposure + futime,
R>                         data = cb_data, time = "futime")
R> fit3 <- fitSmoothHazard(fustat ~ exposure + bs(futime),
R>                         data = cb_data, time = "futime")
R> fit4 <- fitSmoothHazard(fustat ~ exposure*bs(futime),
R>                         data = cb_data, time = "futime")
\end{CodeInput}
\end{CodeChunk}

Note that the fourth model (i.e. \texttt{fit4}) includes an interaction
term between exposure and follow-up time. In other words, this model no
longer exhibit proportional hazards. The evidence of non-proportionality
of hazards in the Stanford Heart Transplant data has been widely
discussed \citep{arjas1988graphical}.

We can then compare the goodness of fit of these four models using the
Akaike Information Criterion (AIC).

\begin{CodeInput}
R> # Compute AIC
R> c("Model1" = AIC(fit1),
R>   "Model2" = AIC(fit2),
R>   "Model3" = AIC(fit3),
R>   "Model4" = AIC(fit4))
\end{CodeInput}
\end{CodeChunk}

As we can, the best fit is the fourth model. By visualizing the hazard
functions for both exposed and unexposed individuals, we can more
clearly see how the hazards are no longer proportional.

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> hazard_data <- expand.grid(exposure = c(0, 1),
R>                            futime = seq(0, 1000,
R>                                         length.out = 100))
R> # Set the offset to zero
R> hazard_data$offset <- 0 
R> # Use predict to get the fitted values, and exponentiate to 
R> # transform to the right scale
R> hazard_data$hazard = exp(predict(fit4, newdata = hazard_data,
R>                                  type = "link"))
R> # Add labels for plots
R> hazard_data$Status = factor(hazard_data$exposure,
R>                             labels = c("NoTrans", "Trans"))
R> 
R> ggplot(hazard_data, aes(futime, hazard, colour = Status)) +
R>     geom_line() +
R>     theme_minimal() +
R>     theme(legend.position = 'top') +
R>     ylab('Hazard') + xlab('Follow-up time')
\end{CodeInput}
\end{CodeChunk}

The non-proportionality seems to be more pronounced at the beginning of
follow-up than the end. Finally, we can turn these estimate of the
hazard function into estimates of the cumulative incidence functions.

\begin{CodeInput}
R> # Compute absolute risk curves
R> newdata <- data.frame(exposure = c(0, 1))
R> absrisk <- absoluteRisk(fit4, newdata = newdata, 
R>                         time = seq(0, 1000, length.out = 100))
R> 
R> colnames(absrisk) <- c("Time", "NoTrans", "Trans")
R> 
R> # Rearrange the data
R> absrisk <- gather(as.data.frame(absrisk),
R>                   "Status", "Risk", -Time)
R>  
R> ggplot(absrisk, aes(Time, Risk, colour = Status)) +
R>   geom_line() +
R>   theme_minimal() +
R>   theme(legend.position = 'top') +
R>   expand_limits(y = 0.5) +
R>   xlab('Follow-up time') + ylab('Cum. Incidence')
\end{CodeInput}
\end{CodeChunk}

Note that we can easily adapt the code above to the situation where a
patient receives a heart transplant at a point in time of interest, for
example after 30 days.

\begin{CodeInput}
R> # Compute hazards---
R> # First, create a list of time points for both exposure status
R> one_yr_haz <- data.frame(futime = seq(0, 365,
R>                                       length.out = 100))
R> one_yr_haz <- mutate(one_yr_haz,
R>                      offset = 0,
R>                      exposure = if_else(futime < 30, 0, 1))
R> one_yr_haz$hazard = exp(predict(fit4, newdata = one_yr_haz,
R>                                 type = "link"))
R> ggplot(one_yr_haz, aes(futime, hazard)) +
R>   geom_line() +
R>   theme_minimal() +
R>   theme(legend.position = 'top') +
R>   ylab('Hazard') + xlab('Follow-up time') +
R>   geom_vline(xintercept = 30, linetype = 'dashed')
\end{CodeInput}
\end{CodeChunk}

We can then compare the 1-year mortality risk without transplant and
with transplant at 30 days.

\begin{CodeInput}
R> absoluteRisk(fit4, newdata = data.frame(exposure = 0),
R>              time = 365)
R> 
R> # Use the trapezoidal rule to estimate the cumulative hazard
R> min_hazard <- min(one_yr_haz$hazard)
R> max_hazard <- max(one_yr_haz$hazard)
R> increment <- 365/99
R> cumhaz_est <- 0.5*increment*(sum(2*one_yr_haz$hazard) - min_hazard - max_hazard)
R> 1 - exp(-cumhaz_est)
\end{CodeInput}
\end{CodeChunk}

As we can see, the risk estimate at 1-year is about 30\% lower if the
patient receives a heart transplant at 30 days.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

In this article, we presented the \proglang{R} package \pkg{casebase}
that provides functions to fit smooth parametric hazards and estimate
cumulative incidence functions using case-base sampling. We outlined the
theoretical underpinnings of the approach, we provided details about our
implementation, and we illustrated the merits of the approach and the
package through three case studies.

\pkg{SmoothHazard}, \pkg{survival} and \pkg{Rstpm2} can handle interval
censoring, whereas \pkg{casebase} currently cannot. Interval censoring
is a simple feature to implement and is a potential improvement to add
to \pkg{casebase}.

\hypertarget{environment-details}{%
\section{Environment Details}\label{environment-details}}

This report was generated on 2020-02-06 17:40:56 using the following
computational environment and dependencies:

\begin{CodeInput}
R> # which R packages and versions?
R> devtools::session_info()
\end{CodeInput}

\begin{CodeOutput}
#> - Session info ---------------------------------------------------------------
#>  setting  value                       
#>  version  R version 3.6.0 (2019-04-26)
#>  os       macOS High Sierra 10.13.6   
#>  system   x86_64, darwin15.6.0        
#>  ui       X11                         
#>  language (EN)                        
#>  collate  en_CA.UTF-8                 
#>  ctype    en_CA.UTF-8                 
#>  tz       America/Montreal            
#>  date     2020-02-06                  
#> 
#> - Packages -------------------------------------------------------------------
#>  package     * version    date       lib
#>  assertthat    0.2.1      2019-03-21 [1]
#>  backports     1.1.5      2019-10-02 [1]
#>  broom         0.5.2      2019-04-07 [1]
#>  callr         3.4.1      2020-01-24 [1]
#>  casebase    * 0.2.1.9001 2020-01-30 [1]
#>  cellranger    1.1.0      2016-07-27 [1]
#>  cli           2.0.1      2020-01-08 [1]
#>  codetools     0.2-16     2018-12-24 [1]
#>  colorspace    1.4-1      2019-03-18 [1]
#>  crayon        1.3.4      2017-09-16 [1]
#>  data.table  * 1.12.8     2019-12-09 [1]
#>  DBI           1.0.0      2018-05-02 [1]
#>  dbplyr        1.4.2      2019-06-17 [1]
#>  desc          1.2.0      2018-05-01 [1]
#>  devtools      2.2.1      2019-09-24 [1]
#>  digest        0.6.23     2019-11-23 [1]
#>  dplyr       * 0.8.3      2019-07-04 [1]
#>  ellipsis      0.3.0      2019-09-20 [1]
#>  evaluate      0.14       2019-05-28 [1]
#>  fansi         0.4.1      2020-01-08 [1]
#>  farver        2.0.3      2020-01-16 [1]
#>  forcats     * 0.4.0      2019-02-17 [1]
#>  foreach       1.4.7      2019-07-27 [1]
#>  fs            1.3.1      2019-05-06 [1]
#>  generics      0.0.2      2018-11-29 [1]
#>  ggplot2     * 3.2.1      2019-08-10 [1]
#>  glmnet      * 3.0-2      2019-12-11 [1]
#>  glue          1.3.1      2019-03-12 [1]
#>  gtable        0.3.0      2019-03-25 [1]
#>  haven         2.2.0      2019-11-08 [1]
#>  hms           0.5.2      2019-10-30 [1]
#>  htmltools     0.4.0      2019-10-04 [1]
#>  httr          1.4.1      2019-08-05 [1]
#>  iterators     1.0.12     2019-07-26 [1]
#>  jsonlite      1.6        2018-12-07 [1]
#>  kableExtra  * 1.1.0      2019-03-16 [1]
#>  knitr         1.27       2020-01-16 [1]
#>  labeling      0.3        2014-08-23 [1]
#>  lattice       0.20-38    2018-11-04 [1]
#>  lazyeval      0.2.2      2019-03-15 [1]
#>  lifecycle     0.1.0      2019-08-01 [1]
#>  lubridate     1.7.4      2018-04-11 [1]
#>  magrittr    * 1.5        2014-11-22 [1]
#>  Matrix      * 1.2-17     2019-03-22 [1]
#>  memoise       1.1.0      2017-04-21 [1]
#>  mgcv          1.8-31     2019-11-09 [1]
#>  modelr        0.1.5      2019-08-08 [1]
#>  munsell       0.5.0      2018-06-12 [1]
#>  nlme          3.1-142    2019-11-07 [1]
#>  pillar        1.4.3      2019-12-20 [1]
#>  pkgbuild      1.0.6      2019-10-09 [1]
#>  pkgconfig     2.0.3      2019-09-22 [1]
#>  pkgload       1.0.2      2018-10-29 [1]
#>  prettyunits   1.1.1      2020-01-24 [1]
#>  processx      3.4.1      2019-07-18 [1]
#>  ps            1.3.0      2018-12-21 [1]
#>  purrr       * 0.3.3      2019-10-18 [1]
#>  R6            2.4.1      2019-11-12 [1]
#>  Rcpp          1.0.3      2019-11-08 [1]
#>  readr       * 1.3.1      2018-12-21 [1]
#>  readxl        1.3.1      2019-03-13 [1]
#>  remotes       2.1.0      2019-06-24 [1]
#>  reprex        0.3.0      2019-05-16 [1]
#>  rlang         0.4.4      2020-01-28 [1]
#>  rmarkdown     2.1        2020-01-20 [1]
#>  rprojroot     1.3-2      2018-01-03 [1]
#>  rstudioapi    0.10       2019-03-19 [1]
#>  rticles       0.12       2019-11-12 [1]
#>  rvest         0.3.5      2019-11-08 [1]
#>  scales        1.1.0      2019-11-18 [1]
#>  sessioninfo   1.1.1      2018-11-05 [1]
#>  shape         1.4.4      2018-02-07 [1]
#>  stringi       1.4.5      2020-01-11 [1]
#>  stringr     * 1.4.0      2019-02-10 [1]
#>  survival    * 3.1-7      2019-11-09 [1]
#>  testthat      2.3.1      2019-12-01 [1]
#>  tibble      * 2.1.3      2019-06-06 [1]
#>  tidyr       * 1.0.0      2019-09-11 [1]
#>  tidyselect    0.2.5      2018-10-11 [1]
#>  tidyverse   * 1.3.0      2019-11-21 [1]
#>  usethis       1.5.1      2019-07-04 [1]
#>  vctrs         0.2.2      2020-01-24 [1]
#>  VGAM          1.1-2      2019-11-21 [1]
#>  viridisLite   0.3.0      2018-02-01 [1]
#>  webshot       0.5.2      2019-11-22 [1]
#>  withr         2.1.2      2018-03-15 [1]
#>  xfun          0.12       2020-01-13 [1]
#>  xml2          1.2.2      2019-08-09 [1]
#>  yaml          2.2.0      2018-07-25 [1]
#>  source                                  
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  Github (sahirbhatnagar/casebase@24359be)
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
#>  CRAN (R 3.6.0)                          
